---
title: 'Reproducible Research: Course Project 2'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Severe weather events and their economic and human health effects  in the United States: 1950 - 2011

## Synopsis
Using data from the NOAA Storm Database, we analyse the economic and human health effects of the observed weather events over the  period 1950 - 2011. While tornadoes have had the most signifant impact on human health, hurricanes and floods have had the greatest economic impact.

## Data Processing

### Read data from compressed csv

```{r, cache=TRUE}
data <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
```

### Parse dates

```{r, message=FALSE}
library(lubridate)
data$BGN_DATE <- mdy_hms(data$BGN_DATE)
```

### Normalize damages

Assume that CROPDMGEXP and PROPDMGEXP express an optional power of 10 multiplier for CROPDMG and PROPDMG respectively.

```{r, cache=TRUE}
multiplier <- function(exp) {
  if (exp == 'k' | exp == 'K') 10**3
  else if (exp == 'm' | exp == 'M') 10**6
  else if (exp == 'b' | exp == 'B') 10**9
  else 1
}

CROPDMG_EXPANDED = 
  with(data,
    CROPDMG * sapply(CROPDMGEXP, multiplier))

PROPDMG_EXPANDED = 
  with(data,
    PROPDMG * sapply(PROPDMGEXP, multiplier))

data$TOTAL_DAMAGES = PROPDMG_EXPANDED + CROPDMG_EXPANDED

```

### Repair erroneous outlier record

Update Record 605943, which is the furthest outlier, and whose REMARKS indicate that the damages were around 70 million.
```{r}
data$TOTAL_DAMAGES[data$REFNUM == 605943] = 70 * 10**6
```

### Normalize EVTYPE factor levels
```{r, message=FALSE}
library(dplyr)

normalize_evtype_level <- function(level) {
  trimws(
  gsub(".*heat.*", "heat wave",
  gsub(".*extreme.*|.*excessive.*|.*extended.*", "extreme/excessive/extended weather", 
  gsub(".*summary.*", "summary",
  gsub(".*record.*", "record weather events", 
  gsub(".*drought.*", "drought",
  gsub(".*sleet.*", "sleet",
  gsub(".*hail.*", "hail",
  gsub(".*snow.*|.*blizzard.*", "snow/blizzard",
  gsub(".*rain.*|.*precipitation.*", "rain/precipitation",
  gsub(".*flood.*|.*fld.*", "flood",
  gsub(".*surf.*|.*swell.*", "surf/swells",
  gsub(".*hurricane.*", "hurricane",
  gsub(".*tornado.*", "tornado",
  gsub(".*volcan.*", "volcanic activity", 
  gsub(".*unseasonabl.*|.*unseasonal.*|.*unusual.*|.*abnormal.*", "unseasonable / abnormal weather", 
  gsub(".*tropical storm.*", "tropical storm",
  gsub(".*land[ ]*slide.*", "landslide",
  gsub(".*mud[ ]*slide.*", "mudslide",
  gsub(".*winter weather.*|.*winter.*mix.*|.*winter storm.*", "winter weather/winter storms",
  gsub(".*thunderstorm.*|.*thundertorm.*|.*thunderstrom.*|.*tunderstorm.*|.*thundersnow.*|.*thuderstorm.*|.*thundeerstorm.*|.*thundertsorm.*|.*thunderstorm.*|.*thunerstorm.*|.thunderestorm.*", "thunderstorm", 
  gsub(".*tstm","thunderstorm", 
  gsub(".*wild[ ]?fire.*|.*wildforest fire.*","wildfire", 
  gsub('mph$','',
  gsub('[0-9()./\\-]',"", 
  tolower(level)
  )))))))))))))))))))))))))
}

normalized_evtype_levels <- sapply(levels(data$EVTYPE), normalize_evtype_level)

normalized_evtype_levels_frame <- data.frame(
  EVTYPE = names(normalized_evtype_levels), 
  EVTYPE_NORMALIZED = normalized_evtype_levels)

rownames(normalized_evtype_levels_frame) <- c() # remove rownames

normalized_data <- left_join(data, normalized_evtype_levels_frame, by=c("EVTYPE"))

# assert that the left join preserved the number of rows...
assertthat::are_equal(dim(data)[1], dim(normalized_data)[1])

# ...and added only 1 column
assertthat::are_equal(dim(data)[2], dim(normalized_data)[2]-1)
```

## Which event has had the greatest impact on human health?

### Fatalities

```{r}
library(ggplot2)

fatalities_by_event <- normalized_data %>%
  group_by(event = EVTYPE_NORMALIZED) %>%
  summarise(cnt = sum(FATALITIES)) %>%
  arrange(desc(cnt))

ggplot(head(fatalities_by_event, 15), aes(x = reorder(event, cnt), y = cnt)) + 
  geom_point() + 
  ggtitle("Deadliest Events\n") + 
  ylab("Fatalities") +
  xlab("") +
  theme(
    axis.text.x=element_text(angle=45,hjust=1,vjust=1),
    plot.title = element_text(hjust = 0.5)
  )
```

Tornadoes have resulted in the greatest amount of fatalities.

### Injuries

```{r}
injuries_by_event <- normalized_data %>%
  group_by(event = EVTYPE_NORMALIZED) %>%
  summarise(cnt = sum(INJURIES)) %>%
  arrange(desc(cnt))

ggplot(head(injuries_by_event, 15), aes(x = reorder(event, cnt), y = cnt)) + 
  geom_point() + 
  ggtitle("Injurious Events\n") + 
  ylab("Injuries") +
  xlab("") +
  theme(
    axis.text.x=element_text(angle=45,hjust=1,vjust=1),
    plot.title = element_text(hjust = 0.5)
  )

```

Tornadoes have resulted in the greatest amount of injuries.

## Which event has had the greatest economic consequences?

```{r}
damages_by_event <- normalized_data %>%
  group_by(event = EVTYPE_NORMALIZED) %>%
  summarise(total_damages_millions = sum(TOTAL_DAMAGES) / 10**9) %>%
  arrange(desc(total_damages_millions))

ggplot(head(damages_by_event, 15), aes(x = reorder(event, total_damages_millions), y = total_damages_millions)) + 
  geom_point() + 
  ggtitle("Economic Impact of Events\n") + 
  ylab("Damages (billions USD)") +
  xlab("") +
  theme(
    axis.text.x=element_text(angle=45,hjust=1,vjust=1),
    plot.title = element_text(hjust = 0.5),
    plot.margin=margin(0,0,0,60)
  )
```

In terms of economic impact, floods and hurricanes have had the greatest impact.

## Results

Tornadoes have had the most signifant impact on human health. Hurricanes and floods have had the greatest economic impact.